---
title: "Lesson 8 — RSM vs truth"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

We have built several methods: brute force, random search, hill climbing, simulated annealing, and RSM. Now comes the **truth reveal** — how close is each method's prediction to the *actual* optimum?

Learning objectives:

1. Load the hidden truth and compare each method's prediction.
2. Compute bias and distance to the true optimum.
3. Create a comparison table and visualisation.
4. Understand why the true optimum lies between grid points.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt_real <- hidden$true_optimum_real
true_yield <- hidden$true_optimum_yield_gL
cat(sprintf("True optimum: pH=%.3f, Temp=%.3f, RPM=%.3f, Inoc=%.3f%%, Sugar=%.3f -> %.3f g/L\n",
            true_opt_real$pH, true_opt_real$Temp_C, true_opt_real$RPM,
            true_opt_real$Inoculum_pct, true_opt_real$Sugar_gL, true_yield))
```

---

# Brute-force best

```{r brute}
cond_means <- df %>%
  group_by(condition_id) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")
bf_best <- cond_means %>% filter(mean_yield == max(mean_yield))
bf_params <- df %>% filter(condition_id == bf_best$condition_id) %>% slice(1)
cat(sprintf("Brute-force best: yield = %.3f g/L\n", bf_best$mean_yield))
```

# RSM prediction

```{r rsm}
factors <- c("pH", "Temp_C", "RPM", "Inoculum_pct", "Sugar_gL")
low <- c(4.0, 25, 50, 5, 20); high <- c(6.0, 35, 200, 15, 60)
names(low) <- names(high) <- factors

df_coded <- df
for (f in factors) df_coded[[paste0(f,"_coded")]] <- 2*(df[[f]]-low[f])/(high[f]-low[f])-1

model <- lm(BC_yield_gL ~ pH_coded + Temp_C_coded + RPM_coded + Inoculum_pct_coded + Sugar_gL_coded +
  pH_coded:Temp_C_coded + Sugar_gL_coded:RPM_coded +
  I(pH_coded^2) + I(Temp_C_coded^2) + I(RPM_coded^2) + I(Inoculum_pct_coded^2) + I(Sugar_gL_coded^2),
  data = df_coded)

b <- coef(model)
b_lin <- c(b["pH_coded"], b["Temp_C_coded"], b["RPM_coded"], b["Inoculum_pct_coded"], b["Sugar_gL_coded"])
b_quad <- c(b["I(pH_coded^2)"], b["I(Temp_C_coded^2)"], b["I(RPM_coded^2)"],
            b["I(Inoculum_pct_coded^2)"], b["I(Sugar_gL_coded^2)"])
A <- diag(2 * b_quad)
A[1,2] <- A[2,1] <- b["pH_coded:Temp_C_coded"]
A[4,5] <- A[5,4] <- b["Sugar_gL_coded:RPM_coded"]
x_star <- -solve(A, b_lin) / 2
rsm_real <- low + (x_star + 1) * (high - low) / 2
rsm_yield <- predict(model, newdata = data.frame(
  pH_coded = x_star[1], Temp_C_coded = x_star[2], RPM_coded = x_star[3],
  Inoculum_pct_coded = x_star[4], Sugar_gL_coded = x_star[5]))

cat(sprintf("RSM optimum: pH=%.3f, Temp=%.3f, RPM=%.3f, Inoc=%.3f, Sugar=%.3f -> %.3f g/L\n",
            rsm_real[1], rsm_real[2], rsm_real[3], rsm_real[4], rsm_real[5], rsm_yield))
```

# Truth reveal: comparison table

```{r comparison, fig.width=10, fig.height=5}
methods <- tibble(
  Method = c("True optimum", "Brute force", "RSM"),
  pH = c(true_opt_real$pH, bf_params$pH, rsm_real[1]),
  Temp_C = c(true_opt_real$Temp_C, bf_params$Temp_C, rsm_real[2]),
  RPM = c(true_opt_real$RPM, bf_params$RPM, rsm_real[3]),
  Inoculum_pct = c(true_opt_real$Inoculum_pct, bf_params$Inoculum_pct, rsm_real[4]),
  Sugar_gL = c(true_opt_real$Sugar_gL, bf_params$Sugar_gL, rsm_real[5]),
  Yield = c(true_yield, bf_best$mean_yield, rsm_yield)
)

# Compute Euclidean distance in real units (normalised)
norm_range <- high - low
methods <- methods %>%
  rowwise() %>%
  mutate(
    Distance = sqrt(sum(((c(pH, Temp_C, RPM, Inoculum_pct, Sugar_gL) -
                          c(true_opt_real$pH, true_opt_real$Temp_C, true_opt_real$RPM,
                            true_opt_real$Inoculum_pct, true_opt_real$Sugar_gL)) / norm_range)^2)),
    Bias = Yield - true_yield
  ) %>%
  ungroup()

print(methods)
```

# Visualise distances

```{r barplot, fig.width=8, fig.height=5}
methods %>%
  filter(Method != "True optimum") %>%
  ggplot(aes(x = Method, y = Distance, fill = Method)) +
  geom_col(alpha = 0.7) +
  labs(x = "", y = "Normalised Euclidean distance to true optimum",
       title = "How close is each method to the truth?") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Exercises

## Exercise 1 — SA and hill climbing comparison
Add SA and hill climbing predictions to the comparison table. Which is closest to the truth?

## Exercise 2 — Why brute force misses
Explain why the brute-force best is not exactly at the true optimum.

## Exercise 3 — Bootstrap the RSM
Fit the RSM 1000 times on bootstrap samples. What is the 95% CI for the predicted optimum pH?

## Exercise 4 — Write a lab report
In 3-4 paragraphs, summarise which method performed best and why. Discuss the trade-offs.

---

# Summary

- The true optimum lies **between** grid points — brute force can't find it exactly.
- RSM interpolates via a statistical model, getting much closer.
- The "aha" moment: model-based approaches beat search-based ones for smooth surfaces.
- **Next:** In Lesson 9, we learn SVR — a non-parametric surrogate.
