---
title: "Lesson 2 — Noise-aware statistics with the BC dataset"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Every measurement carries noise. In Lesson 1 we averaged triplicates; now we learn **why** replication matters and **how to quantify our uncertainty**.

Learning objectives:

1. Compute variance, standard deviation, and standard error by hand and with R functions.
2. Understand why triplicates give more reliable estimates than single runs.
3. Build 95% confidence intervals and interpret them correctly.
4. Compare two conditions with a Welch t-test.
5. Understand the difference between standard deviation and standard error.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
library(car)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
```

---

# Why triplicates beat single runs

## Simulated proof

Let's take one condition and see how the estimated mean stabilises as we add replicates.

```{r simulate-replicates}
set.seed(123)
single_condition <- df %>% filter(condition_id == 1500)
true_mean <- mean(single_condition$BC_yield_gL)

# Simulate: draw 1 to 50 random samples (with replacement) and compute running mean
n_trials <- 50
running_means <- numeric(n_trials)
for (i in 1:n_trials) {
  sample_i <- sample(single_condition$BC_yield_gL, size = i, replace = TRUE)
  running_means[i] <- mean(sample_i)
}

sim_df <- tibble(n = 1:n_trials, mean_yield = running_means)

ggplot(sim_df, aes(x = n, y = mean_yield)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = true_mean, linetype = "dashed", colour = "red") +
  annotate("text", x = n_trials, y = true_mean + 0.05,
           label = sprintf("True mean = %.3f", true_mean), hjust = 1, colour = "red") +
  labs(x = "Number of replicates", y = "Estimated mean yield (g/L)",
       title = "Convergence of the mean with increasing replicates") +
  theme_minimal()
```

---

# Variance, standard deviation, and standard error

## By hand

```{r stats-by-hand}
yields <- c(7.9, 8.2, 8.3)
n <- length(yields)

# Mean
mean_val <- sum(yields) / n
cat(sprintf("Mean = %.4f\n", mean_val))

# Variance (n-1 denominator for sample variance)
variance <- sum((yields - mean_val)^2) / (n - 1)
cat(sprintf("Variance = %.4f\n", variance))

# Standard deviation
sd_val <- sqrt(variance)
cat(sprintf("SD = %.4f\n", sd_val))

# Standard error
se <- sd_val / sqrt(n)
cat(sprintf("SE = %.4f\n", se))
```

## With R functions

```{r stats-r}
cat(sprintf("R mean = %.4f\n", mean(yields)))
cat(sprintf("R var  = %.4f\n", var(yields)))
cat(sprintf("R sd   = %.4f\n", sd(yields)))
cat(sprintf("R SE   = %.4f\n", sd(yields) / sqrt(length(yields))))
```

---

# 95% confidence intervals

The 95% CI for a mean is:

$$\bar{x} \pm t_{n-1,\,0.975} \cdot \frac{s}{\sqrt{n}}$$

```{r confidence-interval}
ci_95 <- function(x) {
  n <- length(x)
  m <- mean(x)
  se <- sd(x) / sqrt(n)
  t_crit <- qt(0.975, df = n - 1)
  lower <- m - t_crit * se
  upper <- m + t_crit * se
  return(list(mean = m, se = se, lower = lower, upper = upper))
}

# Compute CI for condition 1500
cond <- df %>% filter(condition_id == 1500)
ci <- ci_95(cond$BC_yield_gL)
cat(sprintf("Mean = %.3f g/L\n", ci$mean))
cat(sprintf("SE = %.3f g/L\n", ci$se))
cat(sprintf("95%% CI: [%.3f, %.3f] g/L\n", ci$lower, ci$upper))
```

---

# Standard deviation vs standard error

| Measure | What it tells you | Formula |
|---------|-------------------|---------|
| **SD** | How spread out individual measurements are | $\sqrt{\frac{\sum(x_i - \bar{x})^2}{n-1}}$ |
| **SE** | How uncertain the *mean* estimate is | $\frac{s}{\sqrt{n}}$ |
| **CI** | Range where the true mean likely falls | $\bar{x} \pm t \cdot SE$ |

```{r sd-vs-se, fig.width=10, fig.height=5}
cond_means <- df %>%
  group_by(condition_id) %>%
  summarise(
    mean_yield = mean(BC_yield_gL),
    sd_yield   = sd(BC_yield_gL),
    se_yield   = sd(BC_yield_gL) / sqrt(n()),
    pH = first(pH), Temp_C = first(Temp_C),
    .groups = "drop"
  ) %>%
  head(30)

p1 <- ggplot(cond_means, aes(x = reorder(factor(condition_id), mean_yield), y = mean_yield)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_yield - sd_yield, ymax = mean_yield + sd_yield),
                width = 0.3, colour = "steelblue") +
  labs(x = "Condition ID", y = "Yield (g/L)", title = "Mean ± SD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(cond_means, aes(x = reorder(factor(condition_id), mean_yield), y = mean_yield)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_yield - 1.96 * se_yield, ymax = mean_yield + 1.96 * se_yield),
                width = 0.3, colour = "firebrick") +
  labs(x = "Condition ID", y = "Yield (g/L)", title = "Mean ± 95% CI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

library(patchwork)
p1 + p2
```

---

# Comparing two conditions: Welch t-test

```{r welch-t-test}
cond_a <- df %>% filter(condition_id == 1500)
cond_b <- df %>% filter(condition_id == 1600)

cat(sprintf("Condition 1500: mean = %.3f, sd = %.3f\n", mean(cond_a$BC_yield_gL), sd(cond_a$BC_yield_gL)))
cat(sprintf("Condition 1600: mean = %.3f, sd = %.3f\n", mean(cond_b$BC_yield_gL), sd(cond_b$BC_yield_gL)))

test <- t.test(cond_a$BC_yield_gL, cond_b$BC_yield_gL)
cat(sprintf("\nWelch t-test:\n  t = %.3f, df = %.1f, p-value = %.4f\n",
            test$statistic, test$parameter, test$p.value))
cat(sprintf("95%% CI for difference: [%.3f, %.3f]\n", test$conf.int[1], test$conf.int[2]))
```

---

# Exercises

## Exercise 1 — Compute SE for a noisy condition

Pick condition 1 (pH 4.0, 25°C, 50 RPM, 5% inoculum, 20 g/L sugar). Compute its mean, SD, SE, and 95% CI by hand (with a `for` loop), then verify with R functions.

```{r ex1, eval=FALSE}
# Your code here

```

## Exercise 2 — How many replicates for 0.1 g/L precision?

Estimate the number of replicates needed so that the 95% CI half-width is ≤ 0.1 g/L. Use the pooled SD across all conditions as your estimate of σ.

*Hint: solve `t_{n-1} * σ / sqrt(n) = 0.1` for n.*

```{r ex2, eval=FALSE}
# Your code here

```

## Exercise 3 — Multiple comparisons

Compare condition 1500 against every other condition (1 to 3125) using Welch t-tests. How many conditions are significantly different at α = 0.05? Apply a Bonferroni correction.

```{r ex3, eval=FALSE}
# Your code here

```

## Exercise 4 — Error bar plot

Make a bar chart of mean yield for the first 10 pH levels' conditions, with 95% CI error bars. Use `geom_col()` + `geom_errorbar()`.

```{r ex4, eval=FALSE}
# Your code here

```

---

# Summary

- **SD** quantifies measurement variability; **SE** quantifies uncertainty in the mean.
- More replicates → smaller SE → tighter confidence intervals.
- The Welch t-test compares two conditions without assuming equal variances.
- **Next:** In Lesson 3, we learn brute-force search — enumerate all conditions to find the maximum.
