---
title: "Lesson 4 — Random search and grid refinement"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Brute force checks every condition — too expensive in high dimensions. **Random search** samples N random points and picks the best. We then refine by searching the neighbourhood of the best point.

Learning objectives:

1. Implement random search with a budget.
2. Compare random search to brute force on accuracy vs cost.
3. Refine around the best found point with a local grid.
4. Use the "noise-aware" idea: compare means, not single runs.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt <- hidden$true_optimum_real
cat(sprintf("True optimum: pH=%.1f, Temp=%.1f°C, RPM=%.0f, Inoc=%.1f%%, Sugar=%.1f g/L -> %.3f g/L\n",
            true_opt$pH, true_opt$Temp_C, true_opt$RPM, true_opt$Inoculum_pct, true_opt$Sugar_gL,
            hidden$true_optimum_yield_gL))
```

---

# Lookup table

```{r lookup}
# Build lookup: condition_id -> mean yield and parameters
cond_stats <- df %>%
  group_by(condition_id) %>%
  summarise(
    mean_yield = mean(BC_yield_gL),
    sd_yield   = sd(BC_yield_gL),
    pH = first(pH), Temp_C = first(Temp_C), RPM = first(RPM),
    Inoculum_pct = first(Inoculum_pct), Sugar_gL = first(Sugar_gL),
    .groups = "drop"
  )
```

---

# Random search

```{r random-search}
set.seed(42)
budget <- 100  # number of conditions to evaluate

# Randomly sample N condition IDs
random_ids <- sample(cond_stats$condition_id, size = budget, replace = FALSE)
random_results <- cond_stats %>% filter(condition_id %in% random_ids)

best_random <- random_results %>% filter(mean_yield == max(mean_yield))
cat(sprintf("Random search (N=%d): best condition ID = %d, yield = %.3f g/L\n",
            budget, best_random$condition_id, best_random$mean_yield))
cat(sprintf("  Parameters: pH=%.1f, Temp=%.1f, RPM=%.0f, Inoc=%.1f%%, Sugar=%.1f\n",
            best_random$pH, best_random$Temp_C, best_random$RPM,
            best_random$Inoculum_pct, best_random$Sugar_gL))
```

---

# Compare: random search vs brute force

```{r compare, fig.width=8, fig.height=5}
# Run random search multiple times with different budgets
set.seed(42)
n_trials <- 50
budgets <- c(10, 25, 50, 100, 200, 500)

results_list <- list()
for (b in budgets) {
  best_yields <- replicate(n_trials, {
    ids <- sample(cond_stats$condition_id, size = b, replace = FALSE)
    max(cond_stats$mean_yield[cond_stats$condition_id %in% ids])
  })
  results_list[[length(results_list) + 1]] <- tibble(budget = b, best_yield = best_yields)
}
results_df <- bind_rows(results_list)

ggplot(results_df, aes(x = factor(budget), y = best_yield)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = max(cond_stats$mean_yield), linetype = "dashed", colour = "red") +
  annotate("text", x = 1, y = max(cond_stats$mean_yield) + 0.05,
           label = sprintf("Brute-force best = %.3f", max(cond_stats$mean_yield)),
           hjust = 0, colour = "red", size = 3.5) +
  labs(x = "Budget (number of conditions evaluated)",
       y = "Best mean yield found (g/L)",
       title = "Random search: best yield found vs budget") +
  theme_minimal()
```

---

# Grid refinement around the best

```{r refine}
# Find nearest grid points to the best random result
best <- best_random

# Find conditions within ±1 level of each factor
refine_conditions <- cond_stats %>%
  filter(
    abs(pH - best$pH) <= 0.5,
    abs(Temp_C - best$Temp_C) <= 2.5,
    abs(RPM - best$RPM) <= 37.5,
    abs(Inoculum_pct - best$Inoculum_pct) <= 2.5,
    abs(Sugar_gL - best$Sugar_gL) <= 10
  )

cat(sprintf("Refinement grid: %d conditions near the best\n", nrow(refine_conditions)))

refined_best <- refine_conditions %>% filter(mean_yield == max(mean_yield))
cat(sprintf("Refined best: yield = %.3f g/L\n", refined_best$mean_yield))
cat(sprintf("  Parameters: pH=%.1f, Temp=%.1f, RPM=%.0f, Inoc=%.1f%%, Sugar=%.1f\n",
            refined_best$pH, refined_best$Temp_C, refined_best$RPM,
            refined_best$Inoculum_pct, refined_best$Sugar_gL))
```

---

# Noise-aware comparison

When comparing two conditions, always compare **means** with **error bars**, not single runs.

```{r noise-aware, fig.width=8, fig.height=5}
compare_conds <- cond_stats %>%
  filter(condition_id %in% sample(condition_id, 20)) %>%
  arrange(desc(mean_yield)) %>%
  head(10)

ggplot(compare_conds, aes(x = reorder(factor(condition_id), mean_yield), y = mean_yield)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_yield - 1.96 * sd_yield / sqrt(3),
                     ymax = mean_yield + 1.96 * sd_yield / sqrt(3)),
                width = 0.3) +
  labs(x = "Condition ID", y = "Mean yield (g/L)",
       title = "Noise-aware comparison: Mean ± 95% CI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Exercises

## Exercise 1 — Vary the budget

Run random search with budgets of 10, 50, 100, and 500 (100 trials each). How does the distribution of best-found yields change?

## Exercise 2 — Greedy random search

After finding the best of N random points, search all conditions that differ by exactly one level from each factor. Did you improve?

## Exercise 3 — Measure the gap

Compute the gap between the random-search best and the true optimum. Plot this gap as a function of budget.

## Exercise 4 — Multiple restarts

Run random search 10 times with budget=50 and report the best, worst, and median performance.

---

# Summary

- Random search finds near-optimal conditions with far fewer evaluations than brute force.
- Grid refinement around the best point can improve the result.
- Always compare **means** with **error bars** — single runs are unreliable.
- **Next:** In Lesson 5, we learn local search — hill climbing and coordinate ascent.
