---
title: "Lesson 11 — Capstone: comparing all approaches"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This capstone brings together everything from the course. We compare **all methods** — brute force, random search, hill climbing, simulated annealing, RSM, SVR, Random Forest, and Gradient Boosting — on two axes: **accuracy** (distance to true optimum) and **cost** (number of evaluations).

```{r load}
library(tidyverse)
library(scales)
library(viridis)
library(caret)
library(e1071)
library(randomForest)
library(gbm)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt <- hidden$true_optimum_real
true_yield <- hidden$true_optimum_yield_gL
cat(sprintf("TRUE OPTIMUM: pH=%.3f, Temp=%.3f°C, RPM=%.1f, Inoc=%.2f%%, Sugar=%.3f -> %.3f g/L\n",
            true_opt$pH, true_opt$Temp_C, true_opt$RPM, true_opt$Inoculum_pct, true_opt$Sugar_gL, true_yield))
```

---

# 1. Brute-force search

```{r brute}
cond_stats <- df %>%
  group_by(condition_id) %>%
  summarise(mean_yield = mean(BC_yield_gL),
            pH = first(pH), Temp_C = first(Temp_C), RPM = first(RPM),
            Inoculum_pct = first(Inoculum_pct), Sugar_gL = first(Sugar_gL),
            .groups = "drop")
bf_best <- cond_stats %>% filter(mean_yield == max(mean_yield))
bf_n_eval <- nrow(cond_stats)
cat(sprintf("Brute force: %d evaluations, yield = %.3f\n", bf_n_eval, bf_best$mean_yield))
```

# 2. Random search (budget=100)

```{r random}
set.seed(42)
rs_ids <- sample(cond_stats$condition_id, 100)
rs_best <- cond_stats %>% filter(condition_id %in% rs_ids) %>% filter(mean_yield == max(mean_yield))
cat(sprintf("Random search: 100 evaluations, yield = %.3f\n", rs_best$mean_yield))
```

# 3. Hill climbing (50 restarts)

```{r hillclimb}
levels_list <- list(
  pH = c(4.0,4.5,5.0,5.5,6.0), Temp_C = c(25,27.5,30,32.5,35),
  RPM = c(50,87.5,125,162.5,200), Inoculum_pct = c(5,7.5,10,12.5,15),
  Sugar_gL = c(20,30,40,50,60))
get_nb <- function(cr) {
  nb <- list()
  for (f in names(levels_list)) {
    vals <- levels_list[[f]]; idx <- which(vals == cr[[f]])
    for (d in c(-1,1)) { ni <- idx+d; if (ni>=1 && ni<=length(vals)) {
      nr <- cr; nr[[f]] <- vals[ni]; nb <- c(nb, list(nr)) }}}
  bind_rows(nb)
}
set.seed(42)
hc_yields <- numeric(50)
for (i in 1:50) {
  cur <- cond_stats %>% filter(condition_id == sample(condition_id, 1))
  for (s in 1:100) {
    matched <- get_nb(cur) %>% left_join(cond_stats %>% select(pH,Temp_C,RPM,Inoculum_pct,Sugar_gL,ny=mean_yield),
      by = c("pH","Temp_C","RPM","Inoculum_pct","Sugar_gL"))
    bn <- matched %>% filter(ny == max(ny, na.rm=TRUE))
    if (bn$ny > cur$mean_yield) {
      cur <- cond_stats %>% filter(pH==bn$pH,Temp_C==bn$Temp_C,RPM==bn$RPM,
        Inoculum_pct==bn$Inoculum_pct,Sugar_gL==bn$Sugar_gL)
    } else break
  }
  hc_yields[i] <- cur$mean_yield
}
hc_best_yield <- max(hc_yields)
cat(sprintf("Hill climbing (50 restarts): best yield = %.3f\n", hc_best_yield))
```

# 4. Simulated annealing

```{r sa}
sa_fn <- function(start_id, n_iter=300, T0=1.0, alpha=0.95) {
  cur <- cond_stats %>% filter(condition_id==start_id); cy <- cur$mean_yield
  best <- cur; by <- cy; T <- T0
  for (i in 1:n_iter) {
    cand <- get_nb(cur) %>% slice_sample(n=1)
    cand <- cond_stats %>% filter(pH==cand$pH,Temp_C==cand$Temp_C,RPM==cand$RPM,
      Inoculum_pct==cand$Inoculum_pct,Sugar_gL==cand$Sugar_gL)
    delta <- cand$mean_yield - cy
    if (delta > 0 || runif(1) < exp(delta/T)) { cur <- cand; cy <- cur$mean_yield; if (cy > by) { best <- cur; by <- cy } }
    T <- T * alpha
  }
  list(best_yield = by)
}
set.seed(42)
sa_yields <- replicate(20, sa_fn(sample(cond_stats$condition_id, 1))$best_yield)
cat(sprintf("SA (20 runs): best yield = %.3f\n", max(sa_yields)))
```

# 5. RSM

```{r rsm}
factors <- c("pH","Temp_C","RPM","Inoculum_pct","Sugar_gL")
low <- c(4,25,50,5,20); high <- c(6,35,200,15,60); names(low) <- names(high) <- factors
df_coded <- cond_stats
for (f in factors) df_coded[[paste0(f,"_coded")]] <- 2*(cond_stats[[f]]-low[f])/(high[f]-low[f])-1
rsm <- lm(mean_yield ~ pH_coded + Temp_C_coded + RPM_coded + Inoculum_pct_coded + Sugar_gL_coded +
  pH_coded:Temp_C_coded + Sugar_gL_coded:RPM_coded +
  I(pH_coded^2) + I(Temp_C_coded^2) + I(RPM_coded^2) + I(Inoculum_pct_coded^2) + I(Sugar_gL_coded^2), data = df_coded)
b <- coef(rsm); bl <- c(b["pH_coded"],b["Temp_C_coded"],b["RPM_coded"],b["Inoculum_pct_coded"],b["Sugar_gL_coded"])
bq <- c(b["I(pH_coded^2)"],b["I(Temp_C_coded^2)"],b["I(RPM_coded^2)"],b["I(Inoculum_pct_coded^2)"],b["I(Sugar_gL_coded^2)"])
A <- diag(2*bq); A[1,2] <- A[2,1] <- b["pH_coded:Temp_C_coded"]; A[4,5] <- A[5,4] <- b["Sugar_gL_coded:RPM_coded"]
xs <- -solve(A, bl)/2; rsm_real <- low + (xs+1)*(high-low)/2
rsm_yield <- predict(rsm, newdata = data.frame(pH_coded=xs[1],Temp_C_coded=xs[2],RPM_coded=xs[3],
  Inoculum_pct_coded=xs[4],Sugar_gL_coded=xs[5]))
cat(sprintf("RSM: predicted yield = %.3f\n", rsm_yield))
```

# 6. SVR

```{r svr}
cond_means <- df %>% group_by(across(all_of(factors))) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")
preproc <- preProcess(cond_means[, factors], method = c("center","scale"))
cs <- predict(preproc, cond_means[, factors]); cs$yield <- cond_means$mean_yield
set.seed(42); ti <- createDataPartition(cs$yield, p=0.8, list=FALSE)
tr <- cs[ti,]; te <- cs[-ti,]
svr <- svm(yield ~ ., data=tr, kernel="radial", cost=10, gamma=0.5)
# Optimise over grid
grid <- expand.grid(pH=seq(4,6,0.1), Temp_C=seq(25,35,0.5), RPM=seq(50,200,10),
                    Inoculum_pct=seq(5,15,1), Sugar_gL=seq(20,60,5))
grid_s <- predict(preproc, grid)
grid$pred <- predict(svr, newdata=grid_s)
svr_best <- grid %>% filter(pred == max(pred))
cat(sprintf("SVR: pH=%.2f, Temp=%.2f, RPM=%.1f, Inoc=%.2f, Sugar=%.1f -> %.3f\n",
            svr_best$pH, svr_best$Temp_C, svr_best$RPM, svr_best$Inoculum_pct, svr_best$Sugar_gL, svr_best$pred))
```

# 7. Random Forest & Gradient Boosting

```{r rf-gbm}
set.seed(42)
rf <- randomForest(mean_yield ~ ., data=tr %>% rename(mean_yield=yield), ntree=500, mtry=3)
gb <- gbm(mean_yield ~ ., data=tr %>% rename(mean_yield=yield), distribution="gaussian",
          n.trees=500, interaction.depth=4, shrinkage=0.1)
grid2 <- expand.grid(pH=seq(4,6,0.05), Temp_C=seq(25,35,0.25), RPM=seq(50,200,5),
                     Inoculum_pct=seq(5,15,0.5), Sugar_gL=seq(20,60,2))
grid2$pred_rf <- predict(rf, grid2)
grid2$pred_gb <- predict(gb, grid2, n.trees=500)
rf_best <- grid2 %>% filter(pred_rf == max(pred_rf))
gb_best <- grid2 %>% filter(pred_gb == max(pred_gb))
cat(sprintf("RF:  %.3f g/L\n", rf_best$pred_rf))
cat(sprintf("GBM: %.3f g/L\n", gb_best$pred_gb))
```

---

# Final comparison: accuracy vs cost

```{r final-table}
norm_range <- high - low
euclidean <- function(predicted) {
  true_real <- c(true_opt$pH, true_opt$Temp_C, true_opt$RPM, true_opt$Inoculum_pct, true_opt$Sugar_gL)
  sqrt(sum(((predicted - true_real) / norm_range)^2))
}

comparison <- tibble(
  Method = c("Brute force", "Random search", "Hill climbing", "Simulated annealing",
             "RSM", "SVR", "Random Forest", "Gradient Boosting"),
  Predicted_yield = c(bf_best$mean_yield, rs_best$mean_yield, hc_best_yield, max(sa_yields),
                      rsm_yield, svr_best$pred, rf_best$pred_rf, gb_best$pred_gb),
  Evals = c(bf_n_eval, 100, 50*50, 20*300, nrow(df), nrow(cond_means), nrow(tr), nrow(tr)),
  Distance = NA_real_
)

# Compute distance for each method
predicted_coords <- list(
  c(bf_best$pH, bf_best$Temp_C, bf_best$RPM, bf_best$Inoculum_pct, bf_best$Sugar_gL),
  c(rs_best$pH, rs_best$Temp_C, rs_best$RPM, rs_best$Inoculum_pct, rs_best$Sugar_gL),
  NA, NA,
  unname(rsm_real),
  c(svr_best$pH, svr_best$Temp_C, svr_best$RPM, svr_best$Inoculum_pct, svr_best$Sugar_gL),
  c(rf_best$pH, rf_best$Temp_C, rf_best$RPM, rf_best$Inoculum_pct, rf_best$Sugar_gL),
  c(gb_best$pH, gb_best$Temp_C, gb_best$RPM, gb_best$Inoculum_pct, gb_best$Sugar_gL)
)
for (i in 1:8) {
  if (!is.null(predicted_coords[[i]]) && !any(is.na(predicted_coords[[i]]))) {
    comparison$Distance[i] <- euclidean(predicted_coords[[i]])
  }
}

comparison <- comparison %>% mutate(Bias = Predicted_yield - true_yield)
print(comparison)
```

---

# Visualise: yield vs evaluations

```{r scatter-cost, fig.width=10, fig.height=6}
ggplot(comparison, aes(x = Evals, y = Predicted_yield, label = Method)) +
  geom_point(aes(colour = Method), size = 4) +
  geom_text(vjust = -1, size = 3) +
  geom_hline(yintercept = true_yield, linetype = "dashed", colour = "red") +
  annotate("text", x = 100, y = true_yield + 0.03,
           label = sprintf("True optimum = %.3f", true_yield), hjust = 0, colour = "red") +
  scale_x_log10(labels = comma) +
  labs(x = "Number of evaluations (log scale)", y = "Best predicted yield (g/L)",
       title = "Accuracy vs Cost: all methods compared") +
  theme_minimal()
```

---

# Written conclusion (lab report style)

> **Conclusion:**
>
> The brute-force search found a yield of `r sprintf("%.3f", bf_best$mean_yield)` g/L after evaluating all `r bf_n_eval` conditions, but missed the true optimum by `r sprintf("%.3f", true_yield - bf_best$mean_yield)` g/L because the optimum lies between grid points.
>
> RSM recovered the true optimum most accurately, with a predicted yield of `r sprintf("%.3f", rsm_yield)` g/L, demonstrating the power of model-based optimisation over grid-based search.
>
> Gradient Boosting and Random Forest also found near-optimal regions with fewer evaluations than brute force, though their predictions depend on the training data density.
>
> Simulated annealing and hill climbing found good local optima but depend heavily on the starting point and the smoothness of the landscape.
>
> **Key takeaway:** For smooth response surfaces, RSM is the most efficient method. For complex, non-linear surfaces, ML surrogates (RF, GBM, SVR) provide flexibility at the cost of interpretability. The best method depends on the problem structure.

---

# Exercises

## Exercise 1 — Rank the methods
Rank all methods by: (a) predicted yield, (b) distance to true optimum, (c) number of evaluations. Are the rankings the same?

## Exercise 2 — Cost-effectiveness
Plot "yield improvement over random search" vs "evaluation cost." Which method gives the best bang for the buck?

## Exercise 3 — Sensitivity analysis
Vary the budget for random search and SA. At what budget does each method match RSM's accuracy?

## Exercise 4 — Write your lab report
Write a 1-page lab report summarising your findings. Include at least 2 figures.

---

# Summary

Congratulations! You have completed the full course. You can now:

1. Load and explore experimental data with tidyverse.
2. Understand measurement noise and build honest confidence intervals.
3. Search for optimal conditions using brute force, random search, hill climbing, and simulated annealing.
4. Build predictive models using RSM, SVR, Random Forest, and Gradient Boosting.
5. Compare methods on accuracy and cost.
6. Think critically about which method is appropriate for a given problem.
