---
title: "Lesson 10 — Machine Learning surrogates: Random Forest & Gradient Boosting"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

SVR gave us a second surrogate. Now we try **ensemble tree methods** — Random Forest and Gradient Boosting. These build many simple decision trees and combine them for stronger predictions.

Learning objectives:

1. Train Random Forest and Gradient Boosting models with `caret`.
2. Interpret feature importance.
3. Use the surrogate to maximize yield over a continuous search space.
4. Compare ML vs RSM vs SVR.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
library(caret)
library(randomForest)
library(gbm)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt <- hidden$true_optimum_real
true_yield <- hidden$true_optimum_yield_gL
```

---

# Prepare data

```{r prepare}
factors <- c("pH", "Temp_C", "RPM", "Inoculum_pct", "Sugar_gL")
cond_means <- df %>%
  group_by(across(all_of(factors))) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")

set.seed(42)
train_idx <- createDataPartition(cond_means$mean_yield, p = 0.8, list = FALSE)
train <- cond_means[train_idx, ]
test <- cond_means[-train_idx, ]
cat(sprintf("Train: %d, Test: %d\n", nrow(train), nrow(test)))
```

---

# Random Forest

```{r random-forest}
set.seed(42)
rf_model <- randomForest(mean_yield ~ ., data = train, ntree = 500, mtry = 3, importance = TRUE)
pred_rf <- predict(rf_model, newdata = test)
rmse_rf <- sqrt(mean((test$mean_yield - pred_rf)^2))
r2_rf <- cor(test$mean_yield, pred_rf)^2
cat(sprintf("Random Forest: RMSE = %.4f, R² = %.4f\n", rmse_rf, r2_rf))
```

## Feature importance

```{r rf-importance, fig.width=8, fig.height=5}
imp <- as.data.frame(importance(rf_model))
imp$Feature <- rownames(imp)
imp %>%
  arrange(desc(`%IncMSE`)) %>%
  ggplot(aes(x = reorder(Feature, `%IncMSE`), y = `%IncMSE`)) +
  geom_col(fill = "forestgreen", alpha = 0.7) +
  coord_flip() +
  labs(x = "Feature", y = "% Increase in MSE when permuted",
       title = "Random Forest feature importance") +
  theme_minimal()
```

---

# Gradient Boosting

```{r gradient-boosting}
set.seed(42)
gbm_model <- gbm(mean_yield ~ ., data = train, distribution = "gaussian",
                 n.trees = 500, interaction.depth = 4, shrinkage = 0.1, bag.fraction = 0.8)
pred_gbm <- predict(gbm_model, newdata = test, n.trees = 500)
rmse_gbm <- sqrt(mean((test$mean_yield - pred_gbm)^2))
r2_gbm <- cor(test$mean_yield, pred_gbm)^2
cat(sprintf("Gradient Boosting: RMSE = %.4f, R² = %.4f\n", rmse_gbm, r2_gbm))
```

## GBM feature importance

```{r gbm-importance, fig.width=8, fig.height=5}
gbm_imp <- summary(gbm_model, plotit = FALSE)
gbm_imp %>%
  ggplot(aes(x = reorder(var, rel.inf), y = rel.inf)) +
  geom_col(fill = "darkorange", alpha = 0.7) +
  coord_flip() +
  labs(x = "Feature", y = "Relative influence (%)",
       title = "Gradient Boosting feature importance") +
  theme_minimal()
```

---

# Surrogate optimisation

```{r optimise, fig.width=8, fig.height=5}
# Fine grid
grid <- expand.grid(
  pH = seq(4, 6, by = 0.05),
  Temp_C = seq(25, 35, by = 0.25),
  RPM = seq(50, 200, by = 5),
  Inoculum_pct = seq(5, 15, by = 0.5),
  Sugar_gL = seq(20, 60, by = 2)
)

grid$pred_rf <- predict(rf_model, newdata = grid)
grid$pred_gbm <- predict(gbm_model, newdata = grid, n.trees = 500)

best_rf <- grid %>% filter(pred_rf == max(pred_rf))
best_gbm <- grid %>% filter(pred_gbm == max(pred_gbm))

cat(sprintf("RF optimum: pH=%.2f, Temp=%.2f, RPM=%.1f, Inoc=%.2f, Sugar=%.1f -> %.3f g/L\n",
            best_rf$pH, best_rf$Temp_C, best_rf$RPM, best_rf$Inoculum_pct, best_rf$Sugar_gL, best_rf$pred_rf))
cat(sprintf("GBM optimum: pH=%.2f, Temp=%.2f, RPM=%.1f, Inoc=%.2f, Sugar=%.1f -> %.3f g/L\n",
            best_gbm$pH, best_gbm$Temp_C, best_gbm$RPM, best_gbm$Inoculum_pct, best_gbm$Sugar_gL, best_gbm$pred_gbm))
cat(sprintf("True optimum: pH=%.2f, Temp=%.2f, RPM=%.1f, Inoc=%.2f, Sugar=%.1f -> %.3f g/L\n",
            true_opt$pH, true_opt$Temp_C, true_opt$RPM, true_opt$Inoculum_pct, true_opt$Sugar_gL, true_yield))
```

---

# Model comparison

```{r comparison, fig.width=8, fig.height=5}
# Also fit RSM for comparison
df_coded <- cond_means
low <- c(4, 25, 50, 5, 20); high <- c(6, 35, 200, 15, 60)
names(low) <- names(high) <- factors
for (f in factors) df_coded[[paste0(f,"_coded")]] <- 2*(cond_means[[f]]-low[f])/(high[f]-low[f])-1

rsm_model <- lm(mean_yield ~ pH_coded + Temp_C_coded + RPM_coded + Inoculum_pct_coded + Sugar_gL_coded +
  pH_coded:Temp_C_coded + Sugar_gL_coded:RPM_coded +
  I(pH_coded^2) + I(Temp_C_coded^2) + I(RPM_coded^2) + I(Inoculum_pct_coded^2) + I(Sugar_gL_coded^2),
  data = df_coded)

test_coded <- test
for (f in factors) test_coded[[paste0(f,"_coded")]] <- 2*(test[[f]]-low[f])/(high[f]-low[f])-1
pred_rsm <- predict(rsm_model, newdata = test_coded)
rmse_rsm <- sqrt(mean((test$mean_yield - pred_rsm)^2))
r2_rsm <- cor(test$mean_yield, pred_rsm)^2

comparison <- tibble(
  Model = c("RSM", "Random Forest", "Gradient Boosting"),
  RMSE = c(rmse_rsm, rmse_rf, rmse_gbm),
  R2 = c(r2_rsm, r2_rf, r2_gbm)
)
print(comparison)
```

---

# Exercises

## Exercise 1 — Tune Random Forest
Vary `mtry` (1-5) and `ntree` (100-1000). Plot OOB error vs these parameters.

## Exercise 2 — Tune GBM
Vary `shrinkage` (0.01, 0.05, 0.1, 0.2) and `interaction.depth` (2, 4, 6). Which combination gives lowest RMSE?

## Exercise 3 — Observed vs predicted
Make an observed-vs-predicted plot for all three models on the test set.

## Exercise 4 — Learning curve
Plot RMSE vs training set size (10%, 20%, ..., 90%). Do RF and GBM need more data?

---

# Summary

- Random Forest averages many decorrelated trees → robust, low tuning.
- Gradient Boosting builds trees sequentially, correcting errors → often more accurate.
- Feature importance reveals which factors matter most.
- **Next:** In Lesson 11 (capstone), we compare ALL methods head-to-head.
