---
title: "Lesson 1 — R, tidyverse, and plots with the BC dataset"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

You are a bioprocess engineer learning to code. Over this course we will use **R** to answer one question: **what conditions give the maximum bacterial cellulose (BC) yield?**

This lesson gives you the fundamentals you need for everything that follows:

1. **R basics** — variables, vectors, lists, functions, control flow.
2. **tidyverse** — loading and analysing real data tables with `dplyr` and `readr`.
3. **Plotting** — turning numbers into pictures with `ggplot2`.

Work through every cell. When you see **Exercise**, write the code yourself. You will make mistakes — that is the point of dry coding practice.

---

# How to use this notebook (2 minutes)

- Press **Shift+Enter** to run a cell and move to the next one.
- A cell's output appears underneath it.
- New cells: **Insert > R code cell**.
- If you get stuck, read the error message — R tells you the *line* of the problem. Errors are your friend; fix and re-run.

---

# Load the data

```{r load-packages}
library(tidyverse)   # dplyr, ggplot2, readr, tidyr, purrr, stringr
library(scales)
```

```{r load-data}
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
cat(sprintf("Loaded dataset: %d rows x %d columns\n", nrow(df), ncol(df)))
```

---

# Part 1 — R basics (the dry stuff)

Everything in R is a *value* with a *type*. You store values in *variables* and use *functions* to transform them.

## 1.1 Variables and types

```{r variables}
# integers, doubles, strings, logicals
experiment <- 12L          # integer (note the L suffix)
ph         <- 5.5          # double (numeric)
medium     <- "Hestrin"    # character
worked     <- TRUE         # logical

cat(experiment, class(experiment), "\n")
cat(ph, class(ph), "\n")
cat(medium, class(medium), "\n")
cat(worked, class(worked), "\n")
```

**What happened?**

| Step | Code | Variable | Type |
|------|------|----------|------|
| 1 | `experiment <- 12L` | `experiment` | integer |
| 2 | `ph <- 5.5` | `ph` | double |
| 3 | `medium <- "Hestrin"` | `medium` | character |
| 4 | `worked <- TRUE` | `worked` | logical |

In R, the `<-` arrow is the assignment operator (like `=` in Python).

## 1.2 Arithmetic and formatted output

```{r arithmetic}
sugar_gL  <- 50.0
volume_L  <- 2.0
yield_gL  <- 8.1

total_sugar_g    <- sugar_gL * volume_L
yield_per_sugar  <- yield_gL / sugar_gL

cat(sprintf("Total sugar added: %.1f g\n", total_sugar_g))
cat(sprintf("Yield per g of sugar: %.4f g/g\n", yield_per_sugar))
```

## 1.3 Vectors — holding many values

Your three replicates of one condition are a **vector**. R uses 1-based indexing.

```{r vectors}
reps <- c(7.9, 8.2, 8.3)
cat("vector:", reps, "\n")
cat("first replicate (index 1):", reps[1], "\n")
cat("last replicate (index 3):", reps[length(reps)], "\n")
cat("first two (slice 1:2):", reps[1:2], "\n")
cat("length:", length(reps), "\n")
```

**Key difference from Python:** R indices start at **1**, not 0.

## 1.4 Lists — holding mixed types

Lists can hold anything — numbers, strings, even other lists.

```{r lists}
condition <- list(
  condition_id = 2168,
  pH = 5.5,
  Temp_C = 30.0,
  RPM = 87.5,
  Inoculum_pct = 12.5,
  Sugar_gL = 40.0
)

cat(condition$pH, condition$Sugar_gL, "\n")
str(condition)  # shows structure
```

## 1.5 Loops — repeat without copying

A `for` loop visits every element of a vector.

```{r loops}
# Mean by hand
total <- 0
for (y in reps) {
  total <- total + y
}
mean_val <- total / length(reps)
cat(sprintf("mean of %s = %.3f\n", paste(reps, collapse=", "), mean_val))

# Loop with an index
for (i in seq_along(reps)) {
  cat(sprintf("replicate %d: %.2f g/L\n", i, reps[i]))
}
```

**Step-by-step trace of the loop:**

| Iteration | Code | `y` | `total` |
|-----------|------|-----|---------|
| 1 | `total <- 0 + 7.9` | 7.9 | 7.9 |
| 2 | `total <- 7.9 + 8.2` | 8.2 | 16.1 |
| 3 | `total <- 16.1 + 8.3` | 8.3 | 24.4 |

Then `mean_val = 24.4 / 3 = 8.133`.

## 1.6 Functions — package a job

Functions turn a repeated job into one named step. Define once, call many times. **This is the single most useful habit in coding.**

```{r functions}
mean_manual <- function(values) {
  "Return the arithmetic mean of a numeric vector."
  total <- 0
  for (v in values) {
    total <- total + v
  }
  return(total / length(values))
}

cell_yield <- function(sugar_gL, conversion) {
  "Estimate BC yield from sugar and a conversion factor."
  return(sugar_gL * conversion)
}

cat("mean:", mean_manual(reps), "\n")
cat("yield at 50 g/L sugar:", cell_yield(50.0, 0.16), "\n")
```

**Difference between *defining* and *calling*:**

| Expression | What it does |
|-----------|--------------|
| `mean_manual <- function(values) { ... }` | **Defines** the function — stores the recipe |
| `mean_manual(reps)` | **Calls** the function — runs the recipe with `reps` |

## 1.7 Making decisions — if / else / else if

Bioprocessors make decisions (raise pH, slow agitation...). So does code.

```{r if-else}
classify <- function(yield_gL) {
  if (yield_gL >= 8.0) {
    return("excellent")
  } else if (yield_gL >= 5.0) {
    return("good")
  } else {
    return("poor")
  }
}

for (y in c(8.4, 6.1, 2.9)) {
  cat(sprintf("%.1f g/L -> %s\n", y, classify(y)))
}
```

---

# Part 2 — tidyverse: real data in tables

The tidyverse is the workhorse of R data analysis. Its core objects are **tibbles** (data frames) and **verbs** (dplyr functions).

## 2.1 Look at the table

```{r head-tail}
head(df, 3)
```

```{r tail}
tail(df, 3)
```

```{r dim}
cat("Dimensions:", dim(df), "\n")
```

```{r glimpse}
glimpse(df)
```

```{r summary}
summary(df)
```

## 2.2 Filter rows

```{r filter}
high <- df %>% filter(BC_yield_gL >= 7.0)
cat(sprintf("%d runs reached >= 7 g/L = %.2f%% of all runs\n",
            nrow(high), 100 * nrow(high) / nrow(df)))

very_high <- df %>% filter(BC_yield_gL >= 8.0)
cat(sprintf("%d runs reached >= 8 g/L = %.2f%% of all runs\n",
            nrow(very_high), 100 * nrow(very_high) / nrow(df)))
```

## 2.3 Group and summarise

```{r group-summarise}
cond_means <- df %>%
  group_by(condition_id) %>%
  summarise(
    BC_yield_gL = mean(BC_yield_gL),
    pH          = first(pH),
    Temp_C      = first(Temp_C),
    RPM         = first(RPM),
    Inoculum_pct = first(Inoculum_pct),
    Sugar_gL    = first(Sugar_gL),
    .groups = "drop"
  )

cat(sprintf("Unique conditions: %d\n", nrow(cond_means)))
head(cond_means, 3)
```

## 2.4 Summarise by factor

```{r summarise-by-factor}
df %>%
  group_by(pH) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")
```

## 2.5 Two-factor grouping

```{r two-factor}
pivot <- df %>%
  group_by(pH, Temp_C) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop") %>%
  arrange(desc(mean_yield))

cat("Top 5 (pH, Temp) combinations by mean yield:\n")
print(head(pivot, 5))
```

---

# Part 3 — Plotting with ggplot2

`ggplot2` builds plots layer by layer: data → aesthetics → geometry → theme.

## 3.1 Scatter plot: yield vs temperature, coloured by pH

```{r scatter-plot, fig.width=8, fig.height=5}
ggplot(cond_means, aes(x = Temp_C, y = BC_yield_gL, colour = pH)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_colour_viridis_c() +
  labs(
    x = "Temperature (°C)",
    y = "BC yield (g/L)",
    title = "Mean BC yield vs temperature, coloured by pH"
  ) +
  theme_minimal()
```

**What happened?**

| Layer | Code | Purpose |
|-------|------|---------|
| Data | `ggplot(cond_means, ...)` | Start with condition means |
| Aesthetics | `aes(x = Temp_C, y = BC_yield_gL, colour = pH)` | Map columns to visual properties |
| Geometry | `geom_point(size = 2, alpha = 0.7)` | Draw scatter points |
| Scale | `scale_colour_viridis_c()` | Use viridis colour palette |
| Labels | `labs(...)` | Add axis labels and title |
| Theme | `theme_minimal()` | Clean background |

## 3.2 Histogram of yields

```{r histogram, fig.width=8, fig.height=4}
ggplot(df, aes(x = BC_yield_gL)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white", alpha = 0.8) +
  labs(
    x = "BC yield (g/L)",
    y = "Count",
    title = "Distribution of BC yields across all runs"
  ) +
  theme_minimal()
```

## 3.3 Box plot by pH

```{r boxplot, fig.width=8, fig.height=5}
ggplot(df, aes(x = factor(pH), y = BC_yield_gL, fill = factor(pH))) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(
    x = "pH",
    y = "BC yield (g/L)",
    title = "BC yield distribution by pH level"
  ) +
  theme_minimal()
```

## 3.4 Faceted plot: yield vs sugar for each temperature

```{r facets, fig.width=10, fig.height=6}
ggplot(cond_means, aes(x = Sugar_gL, y = BC_yield_gL, colour = factor(pH))) +
  geom_point(alpha = 0.7) +
  facet_wrap(~Temp_C, labeller = label_both) +
  scale_colour_viridis_d() +
  labs(
    x = "Sugar concentration (g/L)",
    y = "BC yield (g/L)",
    title = "Yield vs sugar, faceted by temperature",
    colour = "pH"
  ) +
  theme_minimal()
```

---

# Exercises

## Exercise 1 — Manually average the triplicates (pure R)

Pick `condition_id == 2168`. Filter the dataframe to those three rows, put the three yields into a numeric vector, and compute the mean and the standard deviation **with a `for` loop** (no `mean()` or `sd()`). Then check yourself with `mean()` and `sd()`.

*Hint: standard deviation = sqrt( sum((x - mean)^2) / (n - 1) ).*

```{r ex1, eval=FALSE}
# Your code here

```

---

## Exercise 2 — Boolean filtering

How many individual runs reached a yield of **7 g/L or more**? What fraction of the 9375 runs is that? Then repeat for the stricter threshold `>= 8`.

```{r ex2, eval=FALSE}
# Your code here

```

---

## Exercise 3 — Write a function

Write a function `mean_yield_by_level(factor_name, level)` that returns the mean `BC_yield_gL` for all runs where that factor equals the given level. Call it for every pH level and print the results as a table.

*Hint: `df %>% filter(!!sym(factor_name) == level) %>% pull(BC_yield_gL) %>% mean()`.*

```{r ex3, eval=FALSE}
# Your code here

```

---

## Exercise 4 — Two-factor grouping

Use `group_by` on **both** pH and temperature to compute the mean yield for every (pH, Temp) combination. Which combination gives the highest mean yield? Print the top 5.

```{r ex4, eval=FALSE}
# Your code here

```

---

## Exercise 5 — Colourful scatter plot

Make a scatter plot of **yield vs temperature**, but colour the points by the pH level. Add axis labels, a title, and use the viridis colour scale. Save the figure with `ggsave("my_first_plot.png", width=8, height=5, dpi=150)`.

```{r ex5, eval=FALSE}
# Your code here

```

---

## Exercise 6 (challenge) — The same job with base R only

Re-derive the analysis *without tidyverse*: use `read.csv()` to load `bc_full_factorial_dataset.csv`, accumulate the sum and count of yield per `condition_id` in a named vector (or list), compute the mean, and find the condition with the maximum mean yield. Compare with the tidyverse result.

*Hint: `tapply(df$BC_yield_gL, df$condition_id, mean)` does the group-by in one line.*

```{r ex6, eval=FALSE}
# Your code here

```

---

# Summary

In this lesson you learned:

**R fundamentals:**

- Assignment with `<-`
- Data types: integer, double, character, logical
- Vectors with `c()`, indexing with `[1]`
- Lists with `list()`, accessing with `$`
- `for` loops and `if`/`else`
- Defining and calling functions

**tidyverse:**

- `read_csv()` to load data
- `filter()`, `select()`, `mutate()` to manipulate
- `group_by()` + `summarise()` to aggregate
- `arrange()` to sort
- Piping with `%>%`

**ggplot2:**

- `ggplot(data, aes(...))` + `geom_*()` to build plots
- `scale_colour_viridis_c()` for colour scales
- `facet_wrap()` for small multiples
- `labs()` and `theme_minimal()` for labels and styling

**Next:** In Lesson 2, we learn why triplicates matter — we will explore measurement noise, compute standard errors, and build honest confidence intervals.
