---
title: "Lesson 6 — Global search: simulated annealing"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Hill climbing gets stuck in local optima. **Simulated annealing (SA)** escapes them by accepting worse moves with a probability that decreases over time — like cooling molten metal.

Learning objectives:

1. Implement simulated annealing from scratch.
2. Understand the temperature schedule and acceptance probability.
3. Visualise the SA trajectory and acceptance rate.
4. Compare SA to hill climbing and brute force.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt <- hidden$true_optimum_real

cond_stats <- df %>%
  group_by(condition_id) %>%
  summarise(
    mean_yield = mean(BC_yield_gL),
    pH = first(pH), Temp_C = first(Temp_C), RPM = first(RPM),
    Inoculum_pct = first(Inoculum_pct), Sugar_gL = first(Sugar_gL),
    .groups = "drop"
  )

levels_list <- list(
  pH = c(4.0, 4.5, 5.0, 5.5, 6.0), Temp_C = c(25, 27.5, 30, 32.5, 35),
  RPM = c(50, 87.5, 125, 162.5, 200), Inoculum_pct = c(5, 7.5, 10, 12.5, 15),
  Sugar_gL = c(20, 30, 40, 50, 60)
)

get_neighbours <- function(cond_row) {
  neighbours <- list()
  for (f in names(levels_list)) {
    vals <- levels_list[[f]]
    idx <- which(vals == cond_row[[f]])
    for (delta in c(-1, 1)) {
      new_idx <- idx + delta
      if (new_idx >= 1 && new_idx <= length(vals)) {
        new_row <- cond_row; new_row[[f]] <- vals[new_idx]
        neighbours <- c(neighbours, list(new_row))
      }
    }
  }
  bind_rows(neighbours)
}
```

---

# Simulated annealing from scratch

```{r sa-function}
simulated_annealing <- function(start_id, n_iter = 500, T0 = 1.0, alpha = 0.95) {
  current <- cond_stats %>% filter(condition_id == start_id)
  current_yield <- current$mean_yield
  best <- current
  best_yield <- current_yield

  T <- T0
  history <- tibble(step = 0, yield = current_yield, temperature = T, accepted = TRUE)
  n_accepted <- 0

  for (i in 1:n_iter) {
    # Pick a random neighbour
    neighbours <- get_neighbours(current)
    candidate <- neighbours %>% slice_sample(n = 1)
    candidate <- cond_stats %>%
      filter(pH == candidate$pH, Temp_C == candidate$Temp_C, RPM == candidate$RPM,
             Inoculum_pct == candidate$Inoculum_pct, Sugar_gL == candidate$Sugar_gL)

    delta <- candidate$mean_yield - current_yield

    # Accept if better, or with probability exp(delta / T) if worse
    if (delta > 0 || runif(1) < exp(delta / T)) {
      current <- candidate
      current_yield <- current$mean_yield
      accepted <- TRUE
      n_accepted <- n_accepted + 1
      if (current_yield > best_yield) {
        best <- current
        best_yield <- current_yield
      }
    } else {
      accepted <- FALSE
    }

    # Cool down
    T <- T * alpha
    history <- bind_rows(history, tibble(step = i, yield = current_yield,
                                         temperature = T, accepted = accepted))
  }

  return(list(best = best, best_yield = best_yield, history = history,
              acceptance_rate = n_accepted / n_iter))
}
```

---

# Run simulated annealing

```{r run-sa}
set.seed(42)
start_id <- sample(cond_stats$condition_id, 1)
sa_result <- simulated_annealing(start_id, n_iter = 500, T0 = 1.0, alpha = 0.95)

cat(sprintf("Simulated annealing from condition %d:\n", start_id))
cat(sprintf("  Best yield: %.3f g/L\n", sa_result$best_yield))
cat(sprintf("  Best params: pH=%.1f, Temp=%.1f, RPM=%.0f, Inoc=%.1f%%, Sugar=%.1f\n",
            sa_result$best$pH, sa_result$best$Temp_C, sa_result$best$RPM,
            sa_result$best$Inoculum_pct, sa_result$best$Sugar_gL))
cat(sprintf("  Acceptance rate: %.1f%%\n", 100 * sa_result$acceptance_rate))
```

---

# Visualise the SA trajectory

```{r sa-trajectory, fig.width=10, fig.height=6}
p1 <- ggplot(sa_result$history, aes(x = step, y = yield)) +
  geom_line(colour = "steelblue", linewidth = 0.6) +
  geom_point(aes(colour = accepted), size = 1) +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick")) +
  labs(x = "Iteration", y = "Current yield (g/L)",
       title = "SA trajectory: blue=accepted, red=rejected") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(sa_result$history, aes(x = step, y = temperature)) +
  geom_line(colour = "orange", linewidth = 0.8) +
  labs(x = "Iteration", y = "Temperature",
       title = "Temperature cooling schedule") +
  theme_minimal()

library(patchwork)
p1 / p2
```

---

# Multiple restarts: SA vs Hill Climbing

```{r compare-methods, fig.width=8, fig.height=5}
set.seed(42)
n_restarts <- 30

sa_yields <- numeric(n_restarts)
hc_yields <- numeric(n_restarts)

for (i in 1:n_restarts) {
  start <- sample(cond_stats$condition_id, 1)
  # SA
  sa <- simulated_annealing(start, n_iter = 300, T0 = 1.0, alpha = 0.95)
  sa_yields[i] <- sa$best_yield
  # Hill climb (same start)
  current <- cond_stats %>% filter(condition_id == start)
  for (s in 1:100) {
    matched <- get_neighbours(current) %>%
      left_join(cond_stats %>% select(pH, Temp_C, RPM, Inoculum_pct, Sugar_gL, ny = mean_yield),
                by = c("pH","Temp_C","RPM","Inoculum_pct","Sugar_gL"))
    best_n <- matched %>% filter(ny == max(ny, na.rm = TRUE))
    if (best_n$ny > current$mean_yield) {
      current <- cond_stats %>% filter(pH == best_n$pH, Temp_C == best_n$Temp_C,
        RPM == best_n$RPM, Inoculum_pct == best_n$Inoculum_pct, Sugar_gL == best_n$Sugar_gL)
    } else break
  }
  hc_yields[i] <- current$mean_yield
}

comp <- bind_rows(
  tibble(method = "Simulated Annealing", yield = sa_yields),
  tibble(method = "Hill Climbing", yield = hc_yields)
)

ggplot(comp, aes(x = method, y = yield, fill = method)) +
  geom_boxplot(alpha = 0.6) +
  geom_hline(yintercept = max(cond_stats$mean_yield), linetype = "dashed", colour = "red") +
  annotate("text", x = 1.5, y = max(cond_stats$mean_yield) + 0.03,
           label = sprintf("Brute-force best = %.3f", max(cond_stats$mean_yield)),
           hjust = 0.5, colour = "red", size = 3.5) +
  labs(x = "", y = "Best yield found (g/L)",
       title = "SA vs Hill Climbing: best yield over 30 restarts") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Exercises

## Exercise 1 — Temperature schedule

Try T0 = 0.1, 1.0, and 10.0 with alpha = 0.95. How does the initial temperature affect the search?

## Exercise 2 — Cooling rate

Compare alpha = 0.90, 0.95, and 0.99. Which cools fastest? Which finds the best optimum?

## Exercise 3 — SA trajectory

Plot the SA trajectory for a run that clearly escapes a local optimum. Mark the escape point.

## Exercise 4 — SA + refinement

Run SA, then do grid refinement around the final best. Does it improve?

---

# Summary

- Simulated annealing accepts worse moves with probability exp(Δ/T), allowing escape from local optima.
- The temperature schedule controls exploration vs exploitation.
- SA generally outperforms hill climbing on multi-modal landscapes.
- **Next:** In Lesson 7, we switch from search to **model-based** approaches — response surface methodology (RSM).
