---
title: "Lesson 7 — RSM: multiple linear regression"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Search methods find good points. **Response Surface Methodology (RSM)** builds a statistical *model* of the yield surface, then uses calculus to find the optimum. This is the "aha" moment: we can predict yields we never measured.

Learning objectives:

1. Build a quadratic regression model with `lm()`.
2. Interpret coefficients, R², and ANOVA.
3. Use canonical analysis to find the stationary point.
4. Predict the optimum and visualise the fitted surface.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
library(car)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
```

---

# Code the factors

RSM works best in coded units [-1, +1].

```{r coding}
factors <- c("pH", "Temp_C", "RPM", "Inoculum_pct", "Sugar_gL")
low  <- c(4.0, 25, 50, 5, 20)
high <- c(6.0, 35, 200, 15, 60)
names(low) <- names(high) <- factors

# Create coded columns
df_coded <- df
for (f in factors) {
  df_coded[[paste0(f, "_coded")]] <- 2 * (df[[f]] - low[f]) / (high[f] - low[f]) - 1
}
```

---

# Build the quadratic model

```{r quadratic-model}
# Quadratic formula: y ~ linear + interactions + quadratic
model <- lm(BC_yield_gL ~
  pH_coded + Temp_C_coded + RPM_coded + Inoculum_pct_coded + Sugar_gL_coded +
  pH_coded:Temp_C_coded + Sugar_gL_coded:RPM_coded +
  I(pH_coded^2) + I(Temp_C_coded^2) + I(RPM_coded^2) +
  I(Inoculum_pct_coded^2) + I(Sugar_gL_coded^2),
  data = df_coded
)

summary(model)
```

---

# Interpret the model

## R² and adjusted R²

```{r r-squared}
cat(sprintf("R² = %.4f\n", summary(model)$r.squared))
cat(sprintf("Adjusted R² = %.4f\n", summary(model)$adj.r.squared))
cat(sprintf("Residual SE = %.4f\n", summary(model)$sigma))
```

## Coefficients

```{r coefficients}
coefs <- tidy(model)  # from broom if available, otherwise manual
coefs_table <- data.frame(
  Term = names(coef(model)),
  Estimate = coef(model),
  StdError = summary(model)$coefficients[, 2],
  tValue = summary(model)$coefficients[, 3],
  pValue = summary(model)$coefficients[, 4]
)
print(coefs_table)
```

## ANOVA

```{r anova}
aov_result <- anova(model)
print(aov_result)
```

---

# Find the optimum: canonical analysis

The stationary point of the quadratic is where ∂y/∂x = 0.

```{r canonical}
# Extract coefficients for coded factors
b <- coef(model)
# Linear terms
b_lin <- c(b["pH_coded"], b["Temp_C_coded"], b["RPM_coded"],
           b["Inoculum_pct_coded"], b["Sugar_gL_coded"])

# Quadratic terms (halved, since model uses I(x^2) not 0.5*x^2)
b_quad <- c(b["I(pH_coded^2)"], b["I(Temp_C_coded^2)"], b["I(RPM_coded^2)"],
            b["I(Inoculum_pct_coded^2)"], b["I(Sugar_gL_coded^2)"])

# Interaction terms
b_pH_T <- b["pH_coded:Temp_C_coded"]
b_S_R  <- b["Sugar_gL_coded:RPM_coded"]

# Build Hessian matrix A (2*quadratic coefficients on diagonal, interactions off-diagonal)
A <- matrix(0, 5, 5)
diag(A) <- 2 * b_quad
A[1, 2] <- A[2, 1] <- b_pH_T
A[4, 5] <- A[5, 4] <- b_S_R

# Stationary point: x* = -A^{-1} * b_lin / 2
x_star <- -solve(A, b_lin) / 2

cat("Stationary point (coded):\n")
for (i in 1:5) cat(sprintf("  %s: %.4f\n", factors[i], x_star[i]))

# Convert to real units
real_star <- low + (x_star + 1) * (high - low) / 2
cat("\nStationary point (real units):\n")
for (i in 1:5) cat(sprintf("  %s: %.3f\n", factors[i], real_star[i]))

# Predicted yield at optimum
new_data <- data.frame(
  pH_coded = x_star[1], Temp_C_coded = x_star[2], RPM_coded = x_star[3],
  Inoculum_pct_coded = x_star[4], Sugar_gL_coded = x_star[5]
)
pred_opt <- predict(model, newdata = new_data)
cat(sprintf("\nPredicted optimum yield: %.3f g/L\n", pred_opt))
```

---

# Check if it's a maximum

```{r eigencheck}
eigen_vals <- eigen(A)$values
cat("Eigenvalues of Hessian:\n")
print(eigen_vals)
if (all(eigen_vals < 0)) {
  cat("All eigenvalues negative -> stationary point is a MAXIMUM\n")
} else if (all(eigen_vals > 0)) {
  cat("All eigenvalues positive -> stationary point is a MINIMUM\n")
} else {
  cat("Mixed eigenvalues -> SADDLE POINT (not a true optimum)\n")
}
```

---

# Visualise the fitted surface

```{r fitted-surface, fig.width=10, fig.height=6}
# pH vs Temperature surface
grid <- expand.grid(
  pH_coded = seq(-1, 1, length.out = 50),
  Temp_C_coded = seq(-1, 1, length.out = 50),
  RPM_coded = 0, Inoculum_pct_coded = 0, Sugar_gL_coded = 0
)
grid$pred_yield <- predict(model, newdata = grid)

ggplot(grid, aes(x = pH_coded, y = Temp_C_coded, fill = pred_yield)) +
  geom_tile() +
  scale_fill_viridis_c() +
  geom_point(data = df_coded, aes(x = pH_coded, y = Temp_C_coded),
             inherit.aes = FALSE, size = 0.3, alpha = 0.3) +
  labs(x = "pH (coded)", y = "Temperature (coded)", fill = "Predicted yield (g/L)",
       title = "Fitted response surface: pH vs Temperature") +
  theme_minimal()
```

---

# Exercises

## Exercise 1 — Fit without interactions

Remove the interaction terms. How does R² change? Is the change significant?

## Exercise 2 — Back-transform your model

Write the quadratic model in real (decoded) units. What are the real-unit coefficients?

## Exercise 3 — Residual diagnostics

Plot residuals vs fitted values, QQ plot, and Cook's distance. Are there any concerning patterns?

## Exercise 4 — Predict condition 2000

Use your model to predict the mean yield at condition_id = 2000. Compare with the actual data.

---

# Summary

- RSM fits a quadratic model to all data, then uses calculus to find the optimum.
- `lm()` in R handles quadratic + interaction terms with `I(x^2)` and `:`.
- ANOVA tells us which terms are significant.
- Canonical analysis finds the stationary point.
- **Next:** In Lesson 8, we compare each method's prediction to the true optimum — the "truth reveal."
