---
title: "Lesson 3 — Brute-force search"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Now that we understand the data, we can formalise our question: **what is the optimal operating condition?** The simplest approach is to check every single condition and pick the best one. This is **brute-force search**.

Learning objectives:

1. Formalise the maximisation problem.
2. Build a lookup-table `evaluate()` function with an evaluation counter.
3. Search all 3125 conditions with `which.max`.
4. Visualise the response surface as a heatmap.
5. Understand the "curse of dimensionality."

```{r load}
library(tidyverse)
library(scales)
library(viridis)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
```

---

# The optimisation problem

$$\max_{pH, T, RPM, Ino, Sugar} \; \text{BC yield}(pH, T, RPM, Ino, Sugar)$$

Subject to:
- pH ∈ [4.0, 6.0]
- Temp ∈ [25, 35] °C
- RPM ∈ [50, 200]
- Inoculum ∈ [5, 15] %
- Sugar ∈ [20, 60] g/L

---

# A lookup-table evaluate() function

```{r evaluate}
# Build a lookup from condition_id to mean yield
eval_counter <- 0

evaluate <- function(condition_id) {
  eval_counter <<- eval_counter + 1  # <<- modifies the global variable
  cond <- df %>% filter(condition_id == !!condition_id)
  return(mean(cond$BC_yield_gL))
}
```

---

# Brute-force search

```{r brute-force}
# Get all unique conditions
conditions <- unique(df$condition_id)

# Evaluate every condition
results <- tibble(
  condition_id = conditions,
  mean_yield = sapply(conditions, evaluate)
)

cat(sprintf("Evaluated %d conditions\n", nrow(results)))

# Find the best
best <- results %>% filter(mean_yield == max(mean_yield))
cat(sprintf("Best condition: ID = %d, mean yield = %.3f g/L\n",
            best$condition_id, best$mean_yield))

# Show the parameters of the best condition
best_params <- df %>%
  filter(condition_id == best$condition_id) %>%
  slice(1) %>%
  select(pH, Temp_C, RPM, Inoculum_pct, Sugar_gL)
print(best_params)
```

---

# Curse of dimensionality

How does the cost scale with the number of factors and levels?

```{r curse-of-dim}
L <- 5  # levels per factor
factors <- 1:10
conditions <- L^factors

tibble(Factors = factors, Conditions = conditions) %>%
  ggplot(aes(x = Factors, y = Conditions)) +
  geom_col(fill = "firebrick", alpha = 0.7) +
  scale_y_log10(labels = comma) +
  labs(x = "Number of factors", y = "Number of conditions (log scale)",
       title = "Curse of dimensionality: 5^F grows exponentially") +
  theme_minimal()
```

---

# Visualise the response surface

## Heatmap: pH vs Temperature

```{r heatmap-ph-temp}
heatmap_data <- df %>%
  group_by(pH, Temp_C) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")

ggplot(heatmap_data, aes(x = pH, y = Temp_C, fill = mean_yield)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "pH", y = "Temperature (°C)", fill = "Yield (g/L)",
       title = "Mean BC yield: pH vs Temperature") +
  theme_minimal()
```

## Heatmap: Sugar vs RPM

```{r heatmap-sugar-rpm}
heatmap2 <- df %>%
  group_by(Sugar_gL, RPM) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")

ggplot(heatmap2, aes(x = Sugar_gL, y = RPM, fill = mean_yield)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "Sugar (g/L)", y = "RPM", fill = "Yield (g/L)",
       title = "Mean BC yield: Sugar vs RPM") +
  theme_minimal()
```

---

# Exercises

## Exercise 1 — Find the worst condition

Find the condition with the **lowest** mean yield. What are its parameters?

```{r ex1, eval=FALSE}
# Your code here

```

## Exercise 2 — Cost of brute force

If we had 10 factors with 10 levels each, how many conditions would we need to evaluate? How long would it take at 1 ms per evaluation?

```{r ex2, eval=FALSE}
# Your code here

```

## Exercise 3 — Top 10 conditions

Print the top 10 conditions with their parameters and mean yields.

```{r ex3, eval=FALSE}
# Your code here

```

## Exercise 4 — 3D-ish heatmap

Create a faceted heatmap where each facet is a different (RPM, Inoculum) combination, showing yield as pH vs Sugar.

```{r ex4, eval=FALSE}
# Your code here

```

---

# Summary

- Brute-force search is simple but scales exponentially (curse of dimensionality).
- `which.max()` gives the best condition in one line.
- Heatmaps help visualise multi-factor response surfaces.
- **Next:** In Lesson 4, we learn random search — sampling N random points and refining around the best.
