---
title: "Lesson 5 — Local search: hill climbing & coordinate ascent"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Random search is unbiased but wastes evaluations. **Local search** starts from a point and steps towards better neighbours. Hill climbing takes the steepest improving step; coordinate ascent changes one factor at a time.

Learning objectives:

1. Implement hill climbing from a random start.
2. Implement coordinate ascent (one factor at a time).
3. Understand why local search can get stuck in local optima.
4. Compare multiple random restarts.

```{r load}
library(tidyverse)
library(scales)
library(viridis)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")

# Lookup table: condition_id -> mean yield + parameters
cond_stats <- df %>%
  group_by(condition_id) %>%
  summarise(
    mean_yield = mean(BC_yield_gL),
    pH = first(pH), Temp_C = first(Temp_C), RPM = first(RPM),
    Inoculum_pct = first(Inoculum_pct), Sugar_gL = first(Sugar_gL),
    .groups = "drop"
  )

# Find neighbours (differ by one level in exactly one factor)
levels_list <- list(
  pH = c(4.0, 4.5, 5.0, 5.5, 6.0),
  Temp_C = c(25, 27.5, 30, 32.5, 35),
  RPM = c(50, 87.5, 125, 162.5, 200),
  Inoculum_pct = c(5, 7.5, 10, 12.5, 15),
  Sugar_gL = c(20, 30, 40, 50, 60)
)

get_neighbours <- function(cond_row) {
  # Return all conditions that differ by one level in one factor
  neighbours <- list()
  for (f in names(levels_list)) {
    vals <- levels_list[[f]]
    idx <- which(vals == cond_row[[f]])
    for (delta in c(-1, 1)) {
      new_idx <- idx + delta
      if (new_idx >= 1 && new_idx <= length(vals)) {
        new_row <- cond_row
        new_row[[f]] <- vals[new_idx]
        neighbours <- c(neighbours, list(new_row))
      }
    }
  }
  return(bind_rows(neighbours))
}
```

---

# Hill climbing

Start from a random condition. At each step, evaluate all neighbours and move to the best one (if it improves).

```{r hill-climbing}
set.seed(42)

hill_climb <- function(start_id, max_steps = 100) {
  path <- tibble(step = 0, condition_id = start_id,
                 yield = cond_stats$mean_yield[cond_stats$condition_id == start_id])
  current <- cond_stats %>% filter(condition_id == start_id)

  for (step in 1:max_steps) {
    neighbours <- get_neighbours(current)
    # Match neighbours to cond_stats
    matched <- neighbours %>%
      left_join(cond_stats %>%
                  select(pH, Temp_C, RPM, Inoculum_pct, Sugar_gL, neighbour_yield = mean_yield),
                by = c("pH", "Temp_C", "RPM", "Inoculum_pct", "Sugar_gL"))

    best_neighbour <- matched %>% filter(neighbour_yield == max(neighbour_yield, na.rm = TRUE))

    if (best_neighbour$neighbour_yield > current$mean_yield) {
      current <- cond_stats %>%
        filter(pH == best_neighbour$pH, Temp_C == best_neighbour$Temp_C,
               RPM == best_neighbour$RPM, Inoculum_pct == best_neighbour$Inoculum_pct,
               Sugar_gL == best_neighbour$Sugar_gL)
      path <- bind_rows(path, tibble(step = step, condition_id = current$condition_id,
                                     yield = current$mean_yield))
    } else {
      break  # no improvement -> local optimum
    }
  }
  return(list(path = path, final = current))
}

# Run from a random start
start_id <- sample(cond_stats$condition_id, 1)
result <- hill_climb(start_id)

cat(sprintf("Hill climb from condition %d:\n", start_id))
cat(sprintf("  Steps taken: %d\n", nrow(result$path) - 1))
cat(sprintf("  Final yield: %.3f g/L\n", result$final$mean_yield))
cat(sprintf("  Final params: pH=%.1f, Temp=%.1f, RPM=%.0f, Inoc=%.1f%%, Sugar=%.1f\n",
            result$final$pH, result$final$Temp_C, result$final$RPM,
            result$final$Inoculum_pct, result$final$Sugar_gL))
```

```{r hill-path, fig.width=8, fig.height=4}
ggplot(result$path, aes(x = step, y = yield)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  labs(x = "Step", y = "Mean yield (g/L)",
       title = "Hill climbing trajectory") +
  theme_minimal()
```

---

# Coordinate ascent

Change one factor at a time (cycling through all factors).

```{r coordinate-ascent}
coord_ascent <- function(start_id, max_passes = 20) {
  current <- cond_stats %>% filter(condition_id == start_id)
  best_yield <- current$mean_yield
  path <- tibble(pass = 0, yield = best_yield)

  for (pass in 1:max_passes) {
    improved <- FALSE
    for (f in names(levels_list)) {
      vals <- levels_list[[f]]
      idx <- which(vals == current[[f]])
      for (new_val in vals) {
        if (new_val == current[[f]]) next
        candidate <- cond_stats %>%
          filter(.data[[f]] == new_val,
                 pH == current$pH, Temp_C == current$Temp_C,
                 RPM == current$RPM, Inoculum_pct == current$Inoculum_pct,
                 Sugar_gL == current$Sugar_gL)
        if (nrow(candidate) > 0 && candidate$mean_yield > current$mean_yield) {
          current <- candidate
          best_yield <- current$mean_yield
          improved <- TRUE
        }
      }
    }
    path <- bind_rows(path, tibble(pass = pass, yield = best_yield))
    if (!improved) break
  }

  return(list(path = path, final = current))
}

result_ca <- coord_ascent(start_id)
cat(sprintf("Coordinate ascent from condition %d:\n", start_id))
cat(sprintf("  Passes: %d\n", nrow(result_ca$path) - 1))
cat(sprintf("  Final yield: %.3f g/L\n", result_ca$final$mean_yield))
```

---

# Multiple random restarts

Local search depends on the starting point. Try many restarts.

```{r restarts}
set.seed(42)
n_restarts <- 50
starts <- sample(cond_stats$condition_id, n_restarts)
final_yields <- numeric(n_restarts)

for (i in seq_along(starts)) {
  res <- hill_climb(starts[i])
  final_yields[i] <- res$final$mean_yield
}

cat(sprintf("Best of %d restarts: %.3f g/L\n", n_restarts, max(final_yields)))
cat(sprintf("Worst: %.3f, Median: %.3f\n", min(final_yields), median(final_yields)))

tibble(yield = final_yields) %>%
  ggplot(aes(x = yield)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = max(cond_stats$mean_yield), linetype = "dashed", colour = "red") +
  labs(x = "Final yield (g/L)", y = "Count",
       title = sprintf("Distribution of hill-climb outcomes (%d restarts)", n_restarts)) +
  theme_minimal()
```

---

# Exercises

## Exercise 1 — Compare strategies

Run hill climbing from 50 random starts. How many reach the global optimum (within 0.01 g/L)?

## Exercise 2 — Staircase problem

Plot the trajectory of a hill climb that gets stuck at a local optimum. Annotate where it gets stuck.

## Exercise 3 — Greedy coordinate ascent

Modify coordinate ascent so it picks the **first improving neighbour** instead of the best. How does the result change?

## Exercise 4 — Best vs first

In hill climbing, compare: (a) moving to the **best** neighbour, vs (b) moving to the **first improving** neighbour. Which finds better optima?

---

# Summary

- Hill climbing and coordinate ascent find local optima efficiently.
- Local search depends heavily on the starting point → use multiple restarts.
- The problem with local optima motivates **global search** (next lesson).
- **Next:** In Lesson 6, we learn simulated annealing — a global search method that accepts worse moves with decreasing probability.
