---
title: "Lesson 9 — Support Vector Regression (SVR)"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

RSM assumes a quadratic form. **SVR** is a non-parametric surrogate that can capture more complex patterns. It fits within an "epsilon-tube" around the data.

Learning objectives:

1. Understand why scaling matters for SVR.
2. Train SVR with linear and RBF kernels.
3. Tune hyperparameters (C, epsilon, gamma) with cross-validation.
4. Use SVR as a surrogate to optimise yield.

```{r load}
library(tidyverse)
library(scales)
library(e1071)
library(caret)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
hidden <- jsonlite::fromJSON("../data/hidden_truth.json")
true_opt <- hidden$true_optimum_real
```

---

# Prepare data

```{r prepare}
factors <- c("pH", "Temp_C", "RPM", "Inoculum_pct", "Sugar_gL")
cond_means <- df %>%
  group_by(across(all_of(factors))) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop")

# Scale features (SVR uses distances — scaling is essential)
preproc <- preProcess(cond_means[, factors], method = c("center", "scale"))
cond_scaled <- predict(preproc, cond_means[, factors])
cond_scaled$yield <- cond_means$mean_yield

# Train/test split
set.seed(42)
train_idx <- createDataPartition(cond_scaled$yield, p = 0.8, list = FALSE)
train <- cond_scaled[train_idx, ]
test <- cond_scaled[-train_idx, ]
cat(sprintf("Train: %d, Test: %d\n", nrow(train), nrow(test)))
```

---

# Linear SVR

```{r linear-svr}
svr_linear <- svm(yield ~ ., data = train, kernel = "linear", cost = 1, epsilon = 0.1)
pred_linear <- predict(svr_linear, newdata = test)
rmse_linear <- sqrt(mean((test$yield - pred_linear)^2))
r2_linear <- cor(test$yield, pred_linear)^2
cat(sprintf("Linear SVR: RMSE = %.4f, R² = %.4f\n", rmse_linear, r2_linear))
```

---

# RBF (Radial) SVR

```{r rbf-svr}
svr_rbf <- svm(yield ~ ., data = train, kernel = "radial", cost = 10, gamma = 0.5, epsilon = 0.1)
pred_rbf <- predict(svr_rbf, newdata = test)
rmse_rbf <- sqrt(mean((test$yield - pred_rbf)^2))
r2_rbf <- cor(test$yield, pred_rbf)^2
cat(sprintf("RBF SVR: RMSE = %.4f, R² = %.4f\n", rmse_rbf, r2_rbf))
```

---

# Cross-validated tuning

```{r cv-tune, fig.width=8, fig.height=5}
tuned <- tune.svm(yield ~ ., data = train,
                  gamma = 10^(-3:1), cost = 10^(-1:2),
                  epsilon = 0.05)
cat("Best parameters:\n")
print(tuned$best.parameters)

best_svr <- tuned$best.model
pred_tuned <- predict(best_svr, newdata = test)
rmse_tuned <- sqrt(mean((test$yield - pred_tuned)^2))
r2_tuned <- cor(test$yield, pred_tuned)^2
cat(sprintf("Tuned RBF SVR: RMSE = %.4f, R² = %.4f\n", rmse_tuned, r2_tuned))
```

---

# SVR as surrogate for optimisation

```{r optimise}
# Predict on a fine grid
grid <- expand.grid(
  pH = seq(4, 6, by = 0.1),
  Temp_C = seq(25, 35, by = 0.5),
  RPM = seq(50, 200, by = 10),
  Inoculum_pct = seq(5, 15, by = 1),
  Sugar_gL = seq(20, 60, by = 5)
)

# Scale grid the same way
grid_scaled <- predict(preproc, grid)
grid$pred_yield <- predict(best_svr, newdata = grid_scaled)

best_grid <- grid %>% filter(pred_yield == max(pred_yield))
cat(sprintf("SVR optimum: pH=%.1f, Temp=%.1f, RPM=%.0f, Inoc=%.1f, Sugar=%.0f -> %.3f g/L\n",
            best_grid$pH, best_grid$Temp_C, best_grid$RPM,
            best_grid$Inoculum_pct, best_grid$Sugar_gL, best_grid$pred_yield))
```

---

# Exercises

## Exercise 1 — Compare kernels
Train SVR with linear, polynomial (degree 2, 3), and RBF kernels. Compare RMSE on the test set.

## Exercise 2 — Feature importance
Use permutation importance: shuffle one feature at a time and measure RMSE increase.

## Exercise 3 — Observed vs predicted
Make an observed-vs-predicted scatter plot with the 1:1 line.

## Exercise 4 — SVR vs RSM
Compare SVR and RSM on RMSE and R². Which is better? Why?

---

# Summary

- SVR fits within an epsilon-tube and can capture non-linear patterns.
- **Scaling is essential** — the kernel measures distances.
- Cross-validation tunes hyperparameters without overfitting.
- **Next:** In Lesson 10, we try Random Forest and Gradient Boosting — ensemble ML methods.
