---
title: "Lesson 1 — Solutions (teacher copy)"
author: "Bacterial Cellulose Production Optimization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Give the student `Lesson01_data_exploration.Rmd`. This file has the same six exercises with reference solutions.

Remind the student: many correct solutions exist. Compare approaches, not just answers.

```{r load}
library(tidyverse)
library(scales)
df <- read_csv("../data/bc_full_factorial_dataset.csv", show_col_types = FALSE)
cond_means <- df %>%
  group_by(condition_id) %>%
  summarise(
    BC_yield_gL = mean(BC_yield_gL),
    pH = first(pH), Temp_C = first(Temp_C),
    RPM = first(RPM), Inoculum_pct = first(Inoculum_pct),
    Sugar_gL = first(Sugar_gL),
    .groups = "drop"
  )
```

---

## Exercise 1 — Manually average the triplicates (pure R)

```{r ex1-solution}
rows <- df %>% filter(condition_id == 2168)
yields <- rows$BC_yield_gL
cat("triplicates:", yields, "\n")

# mean by hand
total <- 0
for (y in yields) {
  total <- total + y
}
mean_val <- total / length(yields)

# std by hand
sq <- 0
for (y in yields) {
  sq <- sq + (y - mean_val)^2
}
std <- sqrt(sq / (length(yields) - 1))
cat(sprintf("mean = %.3f  std = %.3f\n", mean_val, std))

# R check
cat(sprintf("R mean = %.3f  R sd = %.3f\n", mean(yields), sd(yields)))
```

---

## Exercise 2 — Boolean filtering

```{r ex2-solution}
high <- df %>% filter(BC_yield_gL >= 7.0)
cat(sprintf("%d runs reached >= 7 g/L = %.2f%% of all runs\n",
            nrow(high), 100 * nrow(high) / nrow(df)))

very_high <- df %>% filter(BC_yield_gL >= 8.0)
cat(sprintf("%d runs reached >= 8 g/L = %.2f%% of all runs\n",
            nrow(very_high), 100 * nrow(very_high) / nrow(df)))
```

---

## Exercise 3 — Write a function

```{r ex3-solution}
mean_yield_by_level <- function(factor_name, level) {
  df %>%
    filter(.data[[factor_name]] == level) %>%
    pull(BC_yield_gL) %>%
    mean()
}

for (level in sort(unique(df$pH))) {
  cat(sprintf("pH %s  ->  %.3f g/L\n", level, mean_yield_by_level("pH", level)))
}
```

---

## Exercise 4 — Two-factor grouping

```{r ex4-solution}
pivot <- df %>%
  group_by(pH, Temp_C) %>%
  summarise(mean_yield = mean(BC_yield_gL), .groups = "drop") %>%
  arrange(desc(mean_yield))

print(head(pivot, 5))
```

---

## Exercise 5 — Colourful scatter plot

```{r ex5-solution, fig.width=8, fig.height=5}
ggplot(cond_means, aes(x = Temp_C, y = BC_yield_gL, colour = pH)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_colour_viridis_c() +
  labs(
    x = "Temperature (°C)",
    y = "BC yield (g/L)",
    title = "Mean BC yield vs temperature, coloured by pH"
  ) +
  theme_minimal()

ggsave("my_first_plot.png", width = 8, height = 5, dpi = 150)
```

---

## Exercise 6 (challenge) — The same job with base R only

```{r ex6-solution}
# Read CSV with base R
df_base <- read.csv("../data/bc_full_factorial_dataset.csv")

# Group-by mean using tapply
means_base <- tapply(df_base$BC_yield_gL, df_base$condition_id, mean)

# Best condition
best_id <- which.max(means_base)
cat(sprintf("best condition (base R): %d, mean = %.3f g/L\n", best_id, means_base[best_id]))

# Compare with tidyverse
pandas_best <- cond_means %>% arrange(desc(BC_yield_gL)) %>% slice(1)
cat(sprintf("best condition (tidyverse): %d, mean = %.3f g/L\n",
            pandas_best$condition_id, pandas_best$BC_yield_gL))
```

---

## Teaching notes

- **Exercise 1** checks loops + the std formula; the R check is a sanity gate for the loop.
- **Exercise 6** is deliberately 'no-tidyverse' — it forces understanding of the group-by logic that `group_by` hides. If the student finds it hard, that is exactly why the exercise exists.
- Expected exercise 2 numbers: `>= 7` is ~16.78% of runs, `>= 8` is ~0.76% (computed from the seed-42 dataset).
